---
title: "Waning Modeling MEthods and Results"
author: "Ivo M. Foppa"
date: "August 17, 2018"
output: pdf_document
fig_caption: yes
header-includes:
- \usepackage{placeins}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`r load('workspace.RData')`

## Model and parameters
We used a simple, unstructured, susceptible-infectious-removed (=SIR) model to simulate the influenza transmission in a large population. The following parameters were chosen:

\begin{table}[ht]
\centering
	\begin{tabular}{|c|c|c|}
	\hline
		Parameter & Symbol & Value \\
		\hline
		Total Pop. & $N$ & $`r formatC(Ntot,format='e',digits=1)`$ \\
		Beginning vaccination uptake & $p_v$ & $`r vacc0`$\\
		vaccination uptake rate& $\nu$ & $`r v`$\\
		Removal rate & $\gamma$ & $`r gamma`$ \\
		Basic reprod. No. & $R_0$ & $`r r0`$\\
		Transmission coeff., unvacc. & $\beta = \frac{R_0\;\gamma}{N}$ & $`r r0`$\\
		Transmission coeff., vacc. & $\beta = \frac{R_0\;\gamma\;(1-\phi)}{N}$ & $`r r0`$\\
		VE & $\phi$ & $0.2,0.3,0.4,0.5$\\
		Pre-esisting immunity & $\epsilon$ & $`r epsilon`$ \\
		\hline
		
	\end{tabular}
\end{table}

The following initial values were used:

\begin{table}[ht]
\centering
	\begin{tabular}{|c|c|c|}
	\hline
		Parameter & Symbol & Value \\
		\hline
		No. susceptible, vacc. & $x_v$ & $`r formatC(xv0,format='e',digits=1)`$ \\
		No. susceptible, unvacc. & $x_{nv}$ & $`r formatC(xnv0,format='e',digits = 1)`$ \\
		No. infectious, vacc. & $y_v=\frac{p_v \;\phi}{p_v \;\phi + 1 - \phi}$ & $`r sapply(1:4,function(x) round(VErnge[x]*vacc0/(VErnge[x]*vacc0 + 1- vacc0),digits = 2))`$ \\
		No. infectious, unvacc. & $y_{nv}$ & $`r round(1 - yv0, digits = 2)`$ \\
		No. removed, vacc. & $z_v$ & $`r 0`$ \\
		No. removed, unvacc. & $z_{nv}$ & $`r 0`$ \\
		\hline
		
	\end{tabular}
\end{table}

The model used is given by the following system or differential equations:
$$\begin{aligned}
      \frac{x_v}{dt} &= - \beta\;(1-\phi)\;x_v\;(y_v + y_{nv}) + \nu\;x_{nv}\\
      \frac{x_{nv}}{dt} &= - \beta\;x_{nv}\;(y_v + y_{nv}) - \nu\;x_{nv} \\
      \frac{y_v}{dt} &= \beta\;(1-\phi)\;x_v\;(y_v + y_{nv}) - y_v\;\gamma \\
      \frac{y_{nv}}{dt} &= \beta\;x_{nv}\;(y_v + y_{nv}) - y_{nv}\;\gamma \\
      \frac{z_v}{dt} &= y_v\;\gamma \\
      \frac{z_{nv}}{dt} &= y_{nv}\;\gamma 
\end{aligned}$$

The system is numerically solved using the `r 'ode'` function from the `r 'deSolve'` R package. 
Test-negative design (TND) studies were simulated by keeping track of the incidence of vaccinated and unvaccinated ``cases'', i.e. new *infecteds*. The observed VE was calculated based on the ratio of the vaccination odds in the cases to the vaccination odds in the population.

## Results

The trajectories are only shown for the periods of time when there was substantial transmission (more than 10 infectious) and aligned at their ``start times''. 

```{r fig1,eval=TRUE,echo=FALSE,fig.cap='Expected VE estimates over time, by VE. \\label{fig:fig1}',fig.pos='H'}
cols <- rainbow(length(VErnge))
plot(times2, VEarr[[1]],type = 'l', col = cols[1], ylim = c(0,max(VErnge)*1.1), ylab = 'VE est.', xlab = 'Day')

for (k in seq_along(VErnge)){
  lines(times2, VEarr[[k]],col = cols[k])
}
legend('bottomright',c('20','30','40','50'), lty = 1, col = cols,bty = 'n' ,title = 'VE (%)')
```

As Figure \ref{fig:fig1} shows, the observed VE always declined, The absulute decline was largest with intermediate VE (VE=30%), while the relative decline was most pronounced for low VE (VE=20%)

\clearpage
```{r,eval=TRUE,fig.pos='ht',echo=FALSE,fig.width=6,fig.height=4}

plot(times2,casearr[[1]],type = 'l', col = cols[1], ylab = 'Incidence', xlab = 'Day')

for (k in seq_along(VErnge)){
  lines(times2,casearr[[k]],col = cols[k])
}
legend('center',c('20','30','40','50'), lty = 1, col = cols,bty = 'n' ,title = 'VE (%)')
```

```{r fig2,eval=TRUE,echo=FALSE,fig.cap='Epi curve (top) and cummulative attack rates (bottom) over time, by VE.\\label{fig:fig2}',fig.pos='ht',fig.width=6,fig.height=4}

plot(times2, cumsum(casearr[[1]])/Ntot,type = 'l', col = cols[1], ylim = c(0,1), ylab = 'Attack rate', xlab = 'Day')

for (k in seq_along(VErnge)){
  lines(times2, cumsum(casearr[[k]])/Ntot,col = cols[k])
}
```

